---
title: "806_final"
author: "Megan Worth"
date: "11/8/2021"
output: pdf_document
editor_options: 
  chunk_output_type: console
---
**Introduction**

**Methods**

**Results**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE}

sup_data <- read.csv("C:/Users/megan/Desktop/806_final/supp_data1.csv", stringsAsFactors = TRUE)

lake_data <- read.csv("C:/Users/megan/Desktop/806_final/lake_data1.csv", stringsAsFactors = TRUE)

otu_data <- read.csv("C:/Users/megan/Desktop/806_final/otu_data1.csv", stringsAsFactors = TRUE)

rel_abun <- read.csv("C:/Users/megan/Desktop/806_final/rel_abun1.csv", stringsAsFactors = TRUE)

```

```{r, echo = FALSE}
library(dplyr)
library(tidyr)
```

```{r, echo = FALSE}

comb1 <- inner_join(otu_data, rel_abun, by = "OTU")
comb2 <- inner_join(sup_data, lake_data)
database <- inner_join(comb1, comb2)


```

# relative abundance of each phlya by pH in 2009
```{r, echo = FALSE}

library(ggplot2)
select(database, Phylum, value, pH, Year) %>%
  filter(Year == 2009, pH > 0) %>% 
  ggplot(mapping = aes(fill = Phylum, x = pH, y = value, color =
  Phylum)) + geom_bar(position="stack", stat="identity") +
  labs(x = "pH", y = "Abundance") +
  theme(legend.text = element_text(size = 6), 
  legend.key.size = unit(0.35, "cm"), 
  legend.key.height = unit(0.35, "cm"),  
  legend.key.width = unit(0.35, "cm")) 

```

# average temp 2007 vs. 2009
```{r, echo = FALSE}

select(database, Month, Year, Temperature) %>%
  group_by(Year, Month) %>%
  filter(Temperature != "NA", Month == "6" | Month == "7" | Month == "8") %>%
  summarize(mean_temp = mean(Temperature)) %>%
  ggplot(mapping = aes(x = Month, y = mean_temp, color = as.factor(Year),
  group = Year, by = Year)) + geom_point() + geom_line() + labs(color = 
  "Year", x = "Month", y = "Temperature (C)") + scale_color_brewer(palette = "Dark2") + scale_x_continuous(breaks=c(6, 7, 8))

```

# dissolved oxygen 2007 vs. 2009
```{r, echo = FALSE}

select(database, Month, Year, DO) %>%
  group_by(Year, Month) %>%
  filter(DO != "NA", Month == "6" | Month == "7" | Month == "8") %>%
  summarize(mean_DO = mean(DO)) %>%
  ggplot(mapping = aes(x = Month, y = mean_DO, color = as.factor(Year),
  group = Year, by = Year)) + geom_point() + geom_line() + labs(color = 
  "Year", x = "Month", y = "Dissolved Oxygen") + scale_color_brewer(palette = "Dark2") + scale_x_continuous(breaks=c(6, 7, 8))

```

# depth vs. dissolved oxygen measured - needs work
```{r, echo = FALSE}

select(database, Depth, DO) %>%
  filter(DO != "NA", Depth != "NA") %>%
  ggplot(mapping = aes(x = Depth, y = DO)) + geom_point() + geom_line() + labs(x = "Depth", y = "Dissolved Oxygen") + scale_color_brewer(palette = "Dark2")

```

# this plot shows abundance of microbes by varying pH, only for 2009 (limited data for 2007)
```{r, echo = FALSE}

select(database, Phylum, value, pH, Year) %>%
  filter(pH > 0, Year == 2009) %>% 
  ggplot(mapping = aes(x = pH, y = value)) + geom_bar(stat="identity") + 
  labs(x = "pH", y = "Abundance") 

```

# Comparison of relative abundances between lake layers in 2009 - needs work
```{r, echo = FALSE}
select(database, Phylum, value, Layer, Year) %>%
  filter(Year == 2009) %>%
  group_by(Layer) %>%
  ggplot(mapping = aes(fill = Phylum, x = Layer, y = value, color =
  Phylum)) + geom_bar(position="stack", stat="identity") +
  labs(x = "Layer", y = "Abundance") +
  theme(legend.text = element_text(size = 6), 
  legend.key.size = unit(0.35, "cm"), 
  legend.key.height = unit(0.35, "cm"),  
  legend.key.width = unit(0.35, "cm")) 
```


**Discussion**



**References**

McMahon, K., S. Jones, A. Shade, R. Newton, E. Read, L. Beversdorf, R. Rohwer, B. Peterson, A. Linz, E. McDaniel, 	G. 	Wolf, and S. Schmitz. 2019. Microbial Observatory at North Temperate Lakes LTER High-resolution temporal and 	spatial dynamics of microbial community structure in freshwater bog lakes 2005 - 2009 original format ver 4. 	Environmental Data Initiative.


